Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-27 paper/ms.tex
Made changes to kernel figure.
Revised Gaussian fit; drafted code for simulating GP in O(N)

A stationary, smooth kernel may generally be expressed as a Taylor series for the $\tau \ne 0$ component,
plus the white noise component:
\begin{equation}
K(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \vert\tau\vert^j.
\end{equation}
Note that this encompasses all smooth, time-staionary kernels, and generally $c_1 \le 0$ as
correlated noise initially decreases with separation in time.  We wish to approximate this
as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
can carry out the approximation in terms of sums of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)}$,
which has a Taylor series expansion:
\begin{equation}
\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} = 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 -
\frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 + \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
\end{equation}
which may be continued to any order in $\tau$ as necessary.
If an approximation to $K(\tau)$ is desired in a Taylor series to some high order $N$, this
requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
chosen to match the coefficients $c_j$.  Not that for $N=1$ we only require a damped exponential
kernel with $\beta_I = 0$.

Another way to view this is in the power spectrum of the Gaussian Process, which is the
Fourier Transform of the autocorrelation function.  The power spectrum of the $K_j$ kernel
is a Lorentzian, for which the frequency is given by $\omega = \beta_{I,j}$, while the
width of the Lorentzian is determined by $\beta_{R,j}$.  The functional form of this
power spectrum is: {\color{red} Check that this is true.}
\begin{equation}
P(\vert f\vert) = a_j \beta_{R,j} \frac{1}{\beta_{R,j}^2 + (2\pi f - \beta_{I,j})^2},
\end{equation}
where $f$ is the frequency in Hz.  The counterpart to the Taylor series is the observation
that a smooth distribution may be approximated arbitrarily precisely by summing together
Lorentzians with various values of $(a_j,\beta_{R,j},\beta_{I,j})$.  In the case that
a peak in the power spectrum is narrow at a specific frequency, then a small value of 
$\beta_{R,j}$ gives a narrow Lorentzian that may approximate that peak.  If the wings
of the distribution decay faster than $f^{-2}$, then differences between Lorentzians
can be arranged to cancel in the wings, giving features in the power spectrum that are
narrower in frequency.

The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels, with various degrees of success:
\begin{itemize}
\item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented with $\beta_R = \beta_I =  0$.
\item {\bf Exponential kernel.} This kernel, $K(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
represented with $\beta_I=0$, $\beta_R = \tau_0^{-1}$.
\item {\bf Cosine kernel.} The cosine kernel, $K(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$, 
may be represented exactly with $\beta_R=0$ and $\beta_I = \frac{2\pi}{P}$.
\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
Nevertheless, it may be well approximated by the sum of three exponential kernels:
\begin{eqnarray}
G(z) &\approx& 2.751807 e^{-1.531250 z} \cos{(0.341489 z)} \\
 &-& 0.354626 e^{-4.753504 z} \cos{(2.042572 z)}\\
 &-& 1.396834 e^{-1.796743 z} \cos{(1.764168 z)},
\end{eqnarray}
which has an accuracy of better than $<4\times 10^{-4}$ for all z.
%[2.751807496353032,1.5312499408182438,-0.34148888180984543,-0.35462563655835033,4.753504292338051,2.0425716395901157,-1.3968344706423566,1.7967431129750422,-1.7641684073374964
\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
\begin{equation}
C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} (\sqrt{(8p+4)}x)^{p-i},
\end{equation} 
where $p$ is a positive integer.  In the limit $p\rightarrow \infty$, this becomes the
squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
values of $p$, the Mat\'ern kernel may be approximated well by the sum of
$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel, while
$p=1$ is relevant to stellar variability, which we describe next.
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ ($p=1$) kernel is given by: 
\begin{equation}
K_{3/2}(\tau) = C_{3/2}(\tau/\tau_0) = (1+v)e^{-v},
\end{equation} 
where $v = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau) \approx (1-b)e^{-v}+be^{-v\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error of the approximation scales as $2/(e^2b) \approx 0.27/b$; for
$b=100$, the error in this approximation is $<0.3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=\textwidth]{kernel_approx.png}
\caption{Approximation of various kernels with the sums of damped cosines with $\tau_0=1$. 
(a). Kernels that may be expressed exactly: constant (dark grey), exponential (light grey), 
cosine (grey), and damped cosine (salmon).
(b). Approximation of the squared-exponential (Gaussian) kernel (pink) with three
damped cosine kernels (magenta). (c). The approximation of the exponential-sine-squared
kernel with $\Gamma =1$ (blue) and $\Gamma = 3$ (green). (d). Two Mat\'ern kernels,
$K_{3/2}$ (yellow/brown) and $K_{5/2}$ (violet).  The Mat\'ern 3/2 kernel
is approximated with $b=100$. }
\label{kernel_approx}
\end{center}
\end{figure}

A standard practice in constructing kernels for solving problems with Gaussian Process
is to add or multiply the preceding standard kernel components.  In the case of damped cosine
kernels, the product of two kernels is: 
\begin{eqnarray}
K_j(\tau)K_k(\tau) &=&
\frac{1}{2} a_k a_j e^{-(\beta_{R,j}+\beta_{R,k}) \tau} \left(\cos{(\beta_{I,j}-\beta_{I,k})\tau}+\cos{(\beta_{I,j}+\beta_{I,k})\tau}\right)\\
&=& \frac{1}{2} \left(K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k}) +K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
\end{eqnarray}
which is the sum of two damped cosine kernels.  Thus, only addition is necessary to
construct new kernels from damped cosine kernels.

The upshot is that the damped cosine is an expressive basis kernel that may be used
to represent a wide range of autocorrelation functions relevant for stellar variability.
We recommend using a series of sums of damped cosines (in which some parameters may be 
fixed to specified values to represent particular kernels) to model different amplitudes
and timescales of variability.  If an initial estimate of the power spectrum or
autocorrelation function is available, then these may be fit with this series to
initialize the coefficients of the damped cosine kernel components.
Since a Gaussian Process
must be positive definite, we recommend modeling the logarithm of $a_j$,
$\beta_{R,j}$, and $\beta_{I,j}$ to enforce this to be the case.  One caveat is
that to produce certain kernels, a negative value of $a_j$ is needed (such
as the Mat\'ern 3/2 kernel).  In this case, the value of the sum of
two coefficients may be sampled from a logarithm, which the individual
values are allowed to be negative;  this will enforce the positive definite
constraint, but may allow for other constraints, such as an autocorrelation
function which has a derivative of zero at $\tau = 0$.

@@

>>>>

A stationary, smooth kernel may generally be expressed as a Taylor series for the $\tau \ne 0$ component,
plus the white noise component:
\begin{equation}
K(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \vert\tau\vert^j.
\end{equation}
Note that this encompasses all smooth, time-stationary kernels, and generally $c_1 \le 0$ as
correlated noise initially declines with separation in time.  We wish to approximate this
as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
can carry out the approximation in terms of sums of damped cosine kernel components
\begin{equation}
K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)},
\end{equation}
which has a Taylor series expansion:
\begin{eqnarray}
\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} &=& 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 \\c &-&
\frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 + \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
\end{eqnarray}
which may be continued to any order in $\tau$ as necessary.
If an approximation to $K(\tau)$ is desired in a Taylor series to some high order $N$, this
requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
chosen to match the coefficients $c_j$.  Note that for $N=1$ we only require a damped exponential
kernel with $\beta_I = 0$.

Another way to view this is in the power spectrum of the gaussian process, which is the
Fourier transform of the autocorrelation function.  The power spectrum of the $K_j$ kernel
is a Lorentzian, for which the frequency is given by $\omega = \beta_{I,j}$, while the
width of the Lorentzian is determined by $\beta_{R,j}$.  The functional form of this
power spectrum is: {\color{red} Check that this is true.}
\begin{equation}
P(\vert f\vert) = a_j \beta_{R,j} \frac{1}{\beta_{R,j}^2 + (2\pi f - \beta_{I,j})^2},
\end{equation}
where $f$ is the frequency in Hz.  The counterpart to the Taylor series is the observation
that a smooth distribution may be approximated arbitrarily precisely by summing together
Lorentzians with various values of $(a_j,\beta_{R,j},\beta_{I,j})$.  In the case that
a peak in the power spectrum is narrow at a specific frequency, then a small value of 
$\beta_{R,j}$ gives a narrow Lorentzian that may approximate that peak.  If the wings
of the distribution decay faster than $f^{-2}$, then differences between Lorentzians
can be arranged to cancel in the wings, giving features in the power spectrum that have
narrower wings.

The functional form of $K_j$ may be used to approximate some commonly used Kernel components.
The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
used to approximate each of these kernels to better than 1\% with various numbers of damped
cosine kernel components:
\begin{itemize}
\item {\bf Constant kernel.} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented exactly with $\beta_R = \beta_I =  0$.
\item {\bf Exponential kernel.} This kernel, $K(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
represented with $\beta_I=0$, $\beta_R = \tau_0^{-1}$.
\item {\bf Cosine kernel.} The cosine kernel, $K(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$, 
may be represented exactly with $\beta_R=0$ and $\beta_I = \frac{2\pi}{P}$.
\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
Likewise, the Fourier transform of the squared exponential kernel is $\hat G(\omega) = e^{-\omega^2/2}$,
which also declines steeply at large $\omega$;  this doesn't match the more gradual decline of a
Lorentzian.  Nevertheless, the squared exponential kernel may be well approximated by the sum of 
three exponential kernels:
\begin{eqnarray}
%G(z) &\approx& 2.751807 e^{-1.531250 z} \cos{(0.341489 z)} \\
% &-& 0.354626 e^{-4.753504 z} \cos{(2.042572 z)}\\
% &-& 1.396834 e^{-1.796743 z} \cos{(1.764168 z)},
G(z) &\approx& 2.670143 e^{-1.500258 z} \cos{(0.346731 z)} \\
 &-& 0.458705 e^{-4.125588 z} \cos{(1.390912 z)}\\
 &-& 1.211438 e^{-1.744602 z} \cos{(1.798436 z)},
\end{eqnarray}
which has an accuracy of better than $<2.5\times 10^{-4}$ for all z.
Note that the last three coefficients depend on the first six as we have constrained
the first three terms of the Taylor series to match that of a Gaussian exactly, $1-z^2/2$.
% [2.6701433354886324,1.5002583976327974,-0.3467305042550739,-0.4587049531317026,4.125587583864702,1.3909117346609636] -1.2114383823569297 1.7446017341309372 1.7984359925567543 
%[2.751807496353032,1.5312499408182438,-0.34148888180984543,-0.35462563655835033,4.753504292338051,2.0425716395901157,-1.3968344706423566,1.7967431129750422,-1.7641684073374964
\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
\begin{equation}
C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} (\sqrt{(8p+4)}x)^{p-i},
\end{equation} 
where $p$ is a positive integer.  In the limit $p\rightarrow \infty$, this becomes the
squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
values of $p$, the Mat\'ern kernel may be approximated well by the sum of
$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel,
$C_{1/2}(x) = \sigma^2 e^{-x}$ ($p=0$), while
$p=1$ is relevant to stellar variability, which we describe next.
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ ($p=1$) kernel is given by: 
\begin{equation}
K_{3/2}(\tau) = C_{3/2}(\tau/\tau_0) = (1+v)e^{-v},
\end{equation} 
where $v = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier transform\footnote{Using the normalization convention: 
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
\begin{equation}
K_{3/2}(\tau) \approx (1-b)e^{-v}+be^{-v\frac{b-1}{b}},
\end{equation} 
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error of the approximation scales as $2/(e^2b) \approx 0.27/b$; for
$b=100$, the error in this approximation is $<0.3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=\textwidth]{kernel_approx.png}
\caption{Approximation of various kernels with the sums of damped cosines.
(a). Kernels that may be expressed exactly: constant (dark grey), exponential (cyan), 
cosine (olive), and damped cosine (salmon).
(b). Approximation of the squared-exponential (Gaussian) kernel (pink) with three
damped cosine kernels (magenta). (c). The approximation of the exponential-sine-squared
kernel with $\Gamma =1$ (blue) and $\Gamma = 3$ (green). (d). Two Mat\'ern kernels,
$K_{3/2}$ (yellow/brown) and $K_{5/2}$ (violet).  The Mat\'ern 3/2 kernel
is approximated with $b=100$. The exact kernels are solid, while approximate are dashed.}
\label{kernel_approx}
\end{center}
\end{figure}

A standard practice in constructing kernels for solving problems with Gaussian processes
is to add or multiply the preceding standard kernel components.  In the case of damped cosine
kernels, the product of two kernels is: 
\begin{eqnarray}
K_j(\tau)K_k(\tau) &=&
\frac{1}{2} a_k a_j e^{-(\beta_{R,j}+\beta_{R,k}) \tau} \left(\cos{(\beta_{I,j}-\beta_{I,k})\tau}+\cos{(\beta_{I,j}+\beta_{I,k})\tau}\right)\\
&=& \frac{1}{2} \left(K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k})\right.\\
&+&\left. K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
\end{eqnarray}
which is the sum of two damped cosine kernels.  Thus, only addition is necessary to
construct new kernels from damped cosine kernels.

The upshot is that the damped cosine is an expressive basis kernel that may be used
to represent a wide range of autocorrelation functions relevant for stellar variability.
We recommend using a series of sums of damped cosines (in which some parameters may be 
fixed to specified values to represent particular kernels) to model different amplitudes
and timescales of variability.  If an initial estimate of the power spectrum or
autocorrelation function is available, then these may be fit with this series to
initialize the coefficients of the damped cosine kernel components.
Since a Gaussian Process
must be positive definite, we recommend modeling the logarithm of $a_j$,
$\beta_{R,j}$, and $\beta_{I,j}$ to enforce this to be the case.  One caveat is
that to produce certain kernels, a negative value of $a_j$ is needed (such
as the Mat\'ern 3/2 kernel).  In this case, the value of the sum of
two coefficients may be sampled from a logarithm, while the individual
values are allowed to be negative;  this will enforce the positive definite
constraint, but may allow for other constraints, such as an autocorrelation
function which has a derivative of zero at $\tau = 0$.

@@

Although we have in mind application of this fast method to stellar variability,
the method is general for one-dimensional GP problems, and may be applied to
other problems.    Within astrophysics, correlated noise (due to the environment,
detector, or modeling uncertainty) may be present in
gravitational wave time series, and so Gaussian processes may be a way to address
this problem \citep{Moore2016}.  Accreting black holes show time series which
may be modeled by correlated noise \citep{Kelly2014};  indeed, this was the
motivation for the original technique developed by Rybicki \& Press \citep{1992ApJ...398..169R,1995PhRvL..74.1060R}.
This approach may be broadly used for characterizing quasar variability \citep{2010ApJ...721.1014M},
measuring time lags with reverberation mapping \citep{2011ApJ...735...80Z},
and modeling time delays in multiply-imaged gravitationally-lensed systems \citep{1998ApJ...507..108P}.
